Theories of deliberated choice

### 2022-11-17 simple.tex
Var

An argumentative theory claims that whenever some of its arguments fail to convince $i$ of choosing $\pga$, it can argue something else.
It is said to be truthful iff this claim is correct.
More precisely, its claims are about paths of the sort $\hist = (\arga, …, \ar_2, …, \ar_3, …)$ where $\ar_2 \in \dgi(\arga)$, $\ar_3 \in \dgi(\ar_2)$, …
The following definition makes this precise by mandating that whenever increasing subsequences of $\hist$, involving arguments $\ar_j$ proposed by the theory, are not convincing, then the theory is able to produce yet another argument in defense of its claim.%
\footnote{Intuition.
If $a_1$, used by the theory, is not decisive, then the theory has an answer, namely, $a_2$, which works at least temporarily.
If $a_1$ used by th is not dec, and $a_2$ used by th is not dec *in the same sequence*, then $a_3$ needs to defend *in the same sequence*.
$(…, a_1)$ NO and $(…, a_1, …, a_2)$ NO then $a_3$ defends it.
$\alpha$ of the form $(\alpha^1, \alpha^2)$ where $a_1$ in end of $α^1$ NO and $a_2$ in end of $α^2$ NO.}
\begin{definition}[Truthfulness]
	\label{def:truth}
	An argumentative theory $(\pga, \arga, \dg)$ for the exposure protocol $(I, \allargs, {\leadsto})$ is \emph{truthful} iff 
	$\forall i \in I, \hist \in \allhist, n \in \N, k \in \intvl{1, \card \hist}^n, \ar \in \allargs^n$:
	\begin{equation}
		\left.\begin{aligned}
			&\ar_1 = \arga \\
			&\forall 1 ≤ j < n: \ar_{j + 1} = \dg(\ar_j, α_{\intvl{1, k_j}}) \in \histbend{k_{j + 1}} \\
			&\forall 1 ≤ j ≤ n: \hist_{\intvl{1, k_j}} \nileadsto \pga
		\end{aligned}\right\}
		⇒ \dg(\ar_n, \hist) ≠ \emptyset.
	\end{equation}
\end{definition}
\begin{axiom}[Answerability]
	\label{ax:ans}
	$\forall \ar, \ar': [\exists \hist \suchthat (\hist, \ar, \ar') \nileadsto \pga \land \exists \hist \suchthat (\hist, \ar, \ar') \ileadsto \pga] ⇒ \exists \hist, \ar_c \suchthat (\hist, \ar, \ar', \ar_c) \nileadsto \pga$.
\end{axiom}
\footnote{$a_1, a_2, a_3$ NOT and $a_1, a_2, a_3, a_4$ YES and $a_1', a_2, a_3, a_4$ NO imply that $a_1, a_2, a_3, a_4, a'_1$ NO.
Generalizing: $\alpha$ NO and $\alpha, a_4$ YES and $\alpha', a_4$ NO imply that $\exists a' \in \alpha' \suchthat \alpha, a_4, a'$ NO.
Let’s try with this too powerful axiom, but then we should weaken it, as it could be that $a_1$ AND $a_4$ TOGETHER are decisive, whence the axiom fails. We should have then some argument $a_1 + a_4$ to replace them, and…?}
\begin{axiom}[Reinstatement]
	\label{ax:r}
	$\forall \hist, \ar_1 \in \gargsi: [(\hist, a_1, a) \nileadsto p \land (\hist, a_1, a, a_2, …) \ileadsto p] ⇒ [(\hist, a_2) \ileadsto p \land \forall \hist': (\hist', a_2, a') \ileadsto p ⇒ (\hist', a_1, a') \nileadsto p]$.
\end{axiom}
\begin{example}[No reinstatement]
	Reinstatement is required. Consider $a_1$ not decisive because of $a$ and then $(a_1, a, a_2) \ileadsto p$. But still $a_2$ not decisive: $a2 \nileadsto p$.
\end{example}
\begin{theorem}[Validity]
	If an argumentative theory $(\pga, \arga, \dg)$ is truthful and $\allargs$ is finite, then \cref{ax:ans,ax:r} imply that the theory is correct.
\end{theorem}
\begin{proof}[Sketch of proof]
	Consider any $i$. We have to show that $\pga \in \Phi_i$.%
	\footnote{Assume $(a_4, \arga, a) \nileadsto \pga$ (thus, $a$ attacks $\arga$). 
	By tf, $(a_4, \arga, a, \ar_2, …) \ileadsto \pga$.
	We get $(a_3, \arga, a, \ar_2) \ileadsto \pga$ by answ.
	By reinst: $(\ar_2)$ YES.}

>>>>

An argumentative theory claims that whenever some of its arguments fail to convince $i$ of choosing $\pga$, it can argue something else.
It is said to be truthful iff this claim is correct.
More precisely, its claims are about paths of the sort $\hist = (\arga, …, \ar_2, …, \ar_3, …)$ where $\ar_2 \in \dgi(\arga)$, $\ar_3 \in \dgi(\ar_2)$, …
The following definition makes this precise by mandating that whenever increasing subsequences of $\hist$, involving arguments $\ar_j$ proposed by the theory, are not convincing, then the theory is able to produce yet another argument in defense of its claim.%
\footnote{Intuition.
If $a_1$, used by the theory, is not decisive, then the theory has an answer, namely, $a_2$, which works at least temporarily.
If $a_1$ used by th is not dec, and $a_2$ used by th is not dec *in the same sequence*, then $a_3$ needs to defend *in the same sequence*.
$(…, a_1)$ NO and $(…, a_1, …, a_2)$ NO then $a_3$ defends it.
$\alpha$ of the form $(\alpha^1, \alpha^2)$ where $a_1$ in end of $α^1$ NO and $a_2$ in end of $α^2$ NO.}
\begin{definition}[Truthfulness]
	\label{def:truth}
	An argumentative theory $(\pga, \arga, \dg)$ for the exposure protocol $(I, \allargs, {\leadsto})$ is \emph{truthful} iff 
	$\forall i \in I, \hist \in \allhist, n \in \N, k \in \intvl{1, \card \hist}^n, \ar \in \allargs^n$:
	\begin{equation}
		\left.\begin{aligned}
			&\ar_1 = \arga \\
			&\forall 1 ≤ j < n: \ar_{j + 1} = \dg(\ar_j, α_{\intvl{1, k_j}}) \in \histbend{k_{j + 1}} \\
			&\forall 1 ≤ j ≤ n: \hist_{\intvl{1, k_j}} \nileadsto \pga
		\end{aligned}\right\}
		⇒ \dg(\ar_n, \hist) ≠ \emptyset.
	\end{equation}
\end{definition}
\begin{axiom}[Answerability]
	\label{ax:ans}
	$\forall i \in I, \hist \in \allhist, \ar \in \gargsi$:
	\begin{equation}
		\left.\begin{aligned}
			&\hist \nileadsto \pga \\
			&(\hist, \ar) \ileadsto \pga \\
			&\forall \ar' \in \allargs: (\hist, \ar, \ar') \ileadsto \pga
		\end{aligned}\right\}
		⇒ \ar \ileadstost \pga.
	\end{equation}
\end{axiom}
\footnote{$a_1, a_2, a_3$ NOT and $a_1, a_2, a_3, a_4$ YES and $a_1', a_2, a_3, a_4$ NO imply that $a_1, a_2, a_3, a_4, a'_1$ NO.
Generalizing: $\alpha$ NO and $\alpha, a_4$ YES and $\alpha', a_4$ NO imply that $\exists a' \in \alpha' \suchthat \alpha, a_4, a'$ NO.
Let’s try with this too powerful axiom, but then we should weaken it, as it could be that $a_1$ AND $a_4$ TOGETHER are decisive, whence the axiom fails. We should have then some argument $a_1 + a_4$ to replace them, and…?
In summary, we want to avoid that $a$ is not decisive but we can’t observe it when starting with some $α$ even though $α$ has also no decisive arguments: if $α$ has no decisive argument, then it must be possible to counter it.

In fact, we might need (a weakening of) the following. If $a_1$ is “decisive” when starting with some $α$ (meaning $(α, a_1, a')$ YES for all arguments $a'$), then $a_1$ is decisive. Hopefully: if $(α, a')$ YES for all arguments $a'$, then some $a \in α$ is decisive.}
\footnote{
	There is a problem. Consider the following may hold for some $i$. $a_1, a_5, a_2$ not convincing, $a_1, a_4$ not convincing, whil $a_1, a_5$ and $a_1, a_4, a_2$ and other variants convincing. Then for all $i$ one of $a_1$ or $a_2$ will appear to be decisive although for those people neither is.

	Problem might be solved by observing that $(a_1, a_5, a_2, a_1)$ convincing means that $a_1$ is decisive after all; and not convincing means that we have observed not truthfulness.

	We want that if the theory is invalid (thus some $i$ is not convinced), then either axiom fails or the theory is experimentally falsifiable.
	In the following example, the theory seems invalid but the axiom holds and the theory can’t seem to be defeated experimentally.

	Imagine $i$ thinks as follows. $a_1$ counter $a_5$; $a_2$ counters $a_4$; but $a_4$ and $a_5$ together make $a_1$ and $a_2$ invalid, so that $(a_4, a_5)$ counter both $a_1$ and $a_2$ (and also $a_1$ and $a_2$ together). Also assume that $i$ has short memory so that $a_1, a_4, a_5$ does not convince ($a_4$ and $a_5$ win) while $a_4, a_5, a_1$ do ($a_1$ wins as $a_4$ is too far away). BUT $a_5, a_4, a_1$ should then not convince.
}

\begin{axiom}[Reinstatement]
	\label{ax:r}
	$\forall i \in I, \hist \in \allhist, \ar_1, \ar_2 \in \gargsi, \ar \in \allargs$:
	\begin{equation}
		\left.\begin{aligned}
			&(\hist, \ar_1, \ar) \nileadsto \pga \\
			&(\hist, \ar_1, \ar, \ar_2) \ileadsto \pga \\
			&(\hist, \ar_1, \ar, \ar_2, \ar) \ileadsto \pga
		\end{aligned}\right\}
		⇒ 
		\left\{\begin{aligned}
			&(\hist, \ar_2) \ileadsto \pga \\
			&(\hist, \ar_2, \ar) \ileadsto \pga \\
			&\forall \ar': (\hist, a_2, a') \nileadsto \pga ⇒ (\hist, a_1, a') \nileadsto \pga.
		\end{aligned}\right.
	\end{equation}
\end{axiom}
\begin{example}[Reinstatement is required]
	Consider $a_1$ not decisive because of $a$ and then $(a_1, a, a_2) \ileadsto p$. But still $a_2$ not decisive: $a2 \nileadsto p$.
\end{example}

\begin{axiom}[Self-sufficiency of positive arguments, version for $\ar$]
	$\forall \ar_1, \ar_2 \in \gargsi, \ar \in \allargs, \hist \in \allhist$:
	\begin{equation}
		\left.\begin{aligned}
			&(\hist, \ar_1, \ar) \nileadsto \pga \\
			&(\hist, \ar_2, \ar) \nileadsto \pga
		\end{aligned}\right\}
		⇒ (\hist, \ar_1, \ar_2, \ar) \nileadsto \pga,
	\end{equation}
	and, $\forall \ar_1, \ar_2 \in \gargsi, \hist \in \allhist$:
	\begin{equation}
		\left.\begin{aligned}
			&(\hist, \ar_1) \nileadsto \pga \\
			&(\hist, \ar_2) \nileadsto \pga
		\end{aligned}\right\}
		⇒ (\hist, \ar_1, \ar_2) \nileadsto \pga.
	\end{equation}
\end{axiom}

\begin{theorem}[Validity]
	If an argumentative theory $(\pga, \arga, \dg)$ is truthful and satisfy \cref{ax:ans,ax:r} and $\allargs$ is finite, then the theory is correct.
\end{theorem}
\begin{proof}
	Consider any $i \in I$.
	We have to show that $\pga \in \Phi_i$.
	Assume depth $3$.
	Assume that $\arga$ and $\ar_2$ are not decisive. Let’s show that $\ar_3$ is decisive.
	Assume that $(a_4, \arga, \ar_2, a) \nileadsto \pga$.
	By TF, $(a_4, \arga, \ar_2, a, \ar_3, ?) \ileadsto \pga$ (where $?$ stands for any argument or nothing).
	By Reinst, $(a_4, \arga, \ar_3, ?) \ileadsto \pga$.
	By Self-s, $(a_4, \ar_3, ?) \ileadsto \pga$.

	Now consider $(\ar_5, \ar_2) \ileadsto \pga$. (Or any such path not ending with $\ar_3$.)
	By Answ, $\exists a \suchthat (\ar_5, \ar_2, a) \nileadsto \pga$.
	We obtain $(\ar_5, \ar_3) \ileadsto \pga$.
\end{proof}
\begin{proof}[Sketch of proof]
	Consider any $i$. We have to show that $\pga \in \Phi_i$.%
	\footnote{Assume $(a_4, \arga, a) \nileadsto \pga$ (thus, $a$ attacks $\arga$). 
	By tf, $(a_4, \arga, a, \ar_2, …) \ileadsto \pga$.
	We get $(a_3, \arga, a, \ar_2) \ileadsto \pga$ by answ.
	By reinst: $(\ar_2)$ YES.}
